---
title: ''
output:
    html_document:
      highlight: pygments
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(error = TRUE, echo = FALSE)
```

```{css echo=FALSE}
# code chunks scrollable
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
```


```{r, echo=FALSE}
# to make the css theme to work, <link></link> tags cannot be added directly 
# as <script></script> tags as below.
# it has to be added using a code chunk with the htmltool functions!!!
css_link = tags$link()
css_link$attribs = list(rel="stylesheet", href="vakata-jstree-3.3.5/dist/themes/default/style.min.css")
css_link
```

```{r, eval=FALSE, echo=FALSE}
# this code chunk is purely for adding comments
# below is to add jQuery and jstree javascripts
```
<script src="https://code.jquery.com/jquery-3.3.1.min.js"></script>
<script src="vakata-jstree-3.3.5/dist/jstree.min.js"></script>

---
# javascript code below is to build the file tree interface
# see this for how to implement opening hyperlink: https://stackoverflow.com/questions/18611317/how-to-get-i-get-leaf-nodes-in-jstree-to-open-their-hyperlink-when-clicked-when
---
<script>
  $(function () {
    // create an instance when the DOM is ready
    $('#jstree').jstree().bind("select_node.jstree", function (e, data) {
     window.open( data.node.a_attr.href, data.node.a_attr.target )
    });
  });
</script>

---
# ADD YOUR DATA ANALYSIS CODE AND MARKUP TEXT BELOW TO EXTEND THIS R MARKDOWN FILE
---

#### Significant genes grouped by pfam target name across multiple organisms

---

```{r message=FALSE}
library(tidyverse)
library(knitr)
tblout_files = strsplit(opt$X_p, ',')[[1]]
common_pfams = read.csv(tblout_files[1])[, 'target_name']
for (f in tblout_files[-1]) {
  common_pfams = intersect(common_pfams, read.csv(f)[, 'target_name'])
}

pfams_df = read.csv(tblout_files[1], stringsAsFactors = FALSE)
pfams_df = pfams_df[pfams_df$target_name %in% common_pfams, c('target_name', 'accession', 'query_name')]
pfams_df = pfams_df %>% group_by(target_name) %>%
  mutate(query_name = paste0(unique(query_name), collapse = ',')) %>%
  distinct() %>%
  arrange(target_name)
first_file_name = tail(strsplit(tblout_files[1], '/')[[1]], 1)
colnames(pfams_df)[3] = first_file_name


for (f in tblout_files[-1]) {
  df = read.csv(f, stringsAsFactors = FALSE)
  df = df[df$target_name %in% common_pfams, c('target_name', 'accession', 'query_name')]
  df = df %>% group_by(target_name) %>%
    mutate(query_name = paste0(unique(query_name), collapse = ',')) %>%
    distinct() %>%
    arrange(target_name)
  df = df[, c("target_name", "query_name"), drop=FALSE]
  header = c("target_name", tail(strsplit(f, '/')[[1]], 1))
  colnames(df) = header
  pfams_df = merge(pfams_df, df[, header], by = "target_name")
}



add_links = function(features) {
  items = strsplit(features, ',')[[1]]
  res = tags$ul(
    lapply(strsplit(items, ','), function(item) {
      item_adj = gsub('_', '', item)
      a(paste0(item, ';'), href=paste0('https://www.hardwoodgenomics.org/feature/', item_adj))
    })
  )
  list(res)
}


table_list = list()
table_list[[1]] = tags$tr(lapply(colnames(pfams_df), tags$th))
for (i in 1:nrow(pfams_df)) {
  target_name_url = tags$a(pfams_df[i, 'target_name'], href=paste0('https://pfam.xfam.org/family/', pfams_df[i, 'target_name']))
  accession_url = tags$a(pfams_df[i, 'accession'], href=paste0('https://pfam.xfam.org/family/', pfams_df[i, 'accession']))
  row = c(list(tags$ul(target_name_url)),
          list(tags$ul(accession_url)),
          lapply(pfams_df[i, -c(1:2)], add_links))
  table_list[[i+1]] = tags$tr(lapply(row, tags$td))
}
tags$table(table_list)
```





