---
title: '[FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/) report'
output:
    html_document:
      highlight: pygments
---


```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(error = TRUE, echo = FALSE)
```

```{css, echo=FALSE}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
```

```{r, echo=FALSE}
# to make the css theme to work, <link></link> tags cannot be added directly 
# as <script></script> tags as below.
# it has to be added using a code chunk with the htmltool functions!!!
css_link = tags$link()
css_link$attribs = list(rel="stylesheet", href="vakata-jstree-3.3.5/dist/themes/default/style.min.css")
css_link
```

```{r, eval=FALSE, echo=FALSE}
# this code chunk is purely for adding comments
# below is to add jQuery and jstree javascripts
```

<script src="vakata-jstree-3.3.5/dist/jstree.min.js"></script>


```{r, eval=FALSE, echo=FALSE}
# this code chunk is purely for adding comments
# javascript code below is to build the file tree interface
# see this for how to implement opening hyperlink: https://stackoverflow.com/questions/18611317/how-to-get-i-get-leaf-nodes-in-jstree-to-open-their-hyperlink-when-clicked-when
```
<script>
  jQuery(function () {
    // create an instance when the DOM is ready
    jQuery('#jstree').jstree().bind("select_node.jstree", function (e, data) {
     window.open( data.node.a_attr.href, data.node.a_attr.target )
    });
  });
</script>


---
# ADD YOUR DATA ANALYSIS CODE AND MARKUP TEXT BELOW TO EXTEND THIS R MARKDOWN FILE
---


# Run FastQC

```{bash}
sh ${TOOL_INSTALL_DIR}/build-and-run-job-scripts.sh
```

```{r echo=FALSE,results='asis'}
# display fastqc job script
cat('```bash\n')
cat(readLines(paste0(Sys.getenv('REPORT_FILES_PATH'), '/job-1-script.sh')), sep = '\n')
cat('\n```')
```

# Fastqc Output Visualization

## Overview

```{r eval=TRUE}
read_1_summary = read.csv(paste0(opt$X_d, '/read_1_fastqc/summary.txt'),
                          stringsAsFactors = FALSE,
                          header = FALSE, sep = '\t')[, 2:1]
read_2_summary = read.csv(paste0(opt$X_d, '/read_2_fastqc/summary.txt'),
                          stringsAsFactors = FALSE,
                          header = FALSE, sep = '\t')[, 1]
combined_summary = data.frame(read_1_summary, read_2_summary, stringsAsFactors = FALSE)
names(combined_summary) = c('MODULE', 'Pre-trimming', 'Post-trimming')
combined_summary[combined_summary == 'FAIL'] = 'FAIL (X)'
combined_summary[combined_summary == 'WARN'] = 'WARN (!)'
DT::datatable(combined_summary)
```

```{r 'function definition', echo=FALSE}
extract_data_module = function(fastqc_data, module_name, header = TRUE, comment.char = "") {
  f = readLines(fastqc_data)
  start_line = grep(module_name, f)
  end_module_lines = grep('END_MODULE', f)
  end_line = end_module_lines[which(end_module_lines > start_line)[1]]
  module_data = f[(start_line+1):(end_line-1)]
  writeLines(module_data, '/tmp/temp.txt')
  read.csv('/tmp/temp.txt', sep = '\t', header = header, comment.char = comment.char)
}
```


### Basic Statistics {.tabset}

#### Before

```{r}
fastqc_data_1 = paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt')
module_name = 'Basic Statistics	pass'

basic_statistics = extract_data_module(fastqc_data_1, module_name)
colnames(basic_statistics) = c('Measure', 'Value')
DT::datatable(basic_statistics)
```

#### After

```{r}
fastqc_data_2 = paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt')
module_name = 'Basic Statistics	pass'

basic_statistics = extract_data_module(fastqc_data_2, module_name)
colnames(basic_statistics) = c('Measure', 'Value')
DT::datatable(basic_statistics)
```


### Per base sequence quality

```{r 'per base sequence quality'}
## reads 1
pbsq_1 = extract_data_module(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt'), 'Per base sequence quality')
pbsq_1$id = 1:length(pbsq_1$X.Base)
pbsq_1$trim = 'before'

## reads 2
pbsq_2 = extract_data_module(paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt'), 'Per base sequence quality')
pbsq_2$id = 1:length(pbsq_2$X.Base)
pbsq_2$trim = 'after'

comb_pbsq = rbind(pbsq_1, pbsq_2)
comb_pbsq$trim = factor(levels = c('before', 'after'), comb_pbsq$trim)

p = ggplot(data = comb_pbsq) +
  geom_boxplot(mapping = aes(x = id, 
                             lower = Lower.Quartile, 
                             upper = Upper.Quartile, 
                             middle = Median, 
                             ymin = X10th.Percentile, 
                             ymax = X90th.Percentile,
                             fill = "yellow"),
               stat = 'identity') +
  geom_line(mapping = aes(x = id, y = Mean, color = "red")) +
  scale_x_continuous(name = '\nPosition in read (bp)', breaks = pbsq_2$id, labels = pbsq_2$X.Base) +
  scale_y_continuous(limits = c(0, max(comb_pbsq$Upper.Quartile) + 5)) +
  scale_fill_identity() +
  scale_color_identity() + 
  facet_grid(. ~ trim) +
  theme(axis.text.x = element_text(size = 5),
        panel.background = element_rect(fill = NA),
        panel.grid.major.y = element_line(color = 'blue', size = 0.1))
p
```


### Per tile sequence quality

```{r 'per tile sequence quality'}
## check if 'per tile sequence quality' module exits or not
check_ptsq = grep('Per tile sequence quality', readLines(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt')))
if (length(check_ptsq) > 0) {
    ## reads 1
  ptsq_1 = extract_data_module(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt'), 'Per tile sequence quality')
  ptsq_1$trim = 'before'
  
  ## reads 2
  ptsq_2 = extract_data_module(paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt'), 'Per tile sequence quality')
  ptsq_2$trim = 'after'
  
  comb_ptsq = rbind(ptsq_1, ptsq_2)
  comb_ptsq$trim = factor(levels = c('before', 'after'), comb_ptsq$trim)
  comb_ptsq$Base = factor(levels = unique(comb_ptsq$Base), comb_ptsq$Base)
  
  # convert integers to charaters
  # comb_ptsq$Tile = as.character(comb_ptsq$X.Tile)
  
  p = ggplot(data = comb_ptsq) +
    geom_raster(mapping = aes(x = Base, y = X.Tile, fill = Mean)) + 
    facet_grid(. ~ trim) + 
    scale_x_discrete(name = "\nPosition in read (bp)") +
    scale_y_continuous(name = "") +
    scale_fill_gradient(low = "blue", high = "red") +
    theme(axis.text.x = element_text(size = 5, angle = 90),
          axis.text.y = element_text(size = 5),
        panel.background = element_rect(fill = NA))
  ggplotly(p)
} else {
  print('No "per tile sequence quality" data')
}
```

### Per sequence quality score

```{r 'Per sequence quality score'}
## reads 1
psqs_1 = extract_data_module(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt'), 'Per sequence quality scores')
psqs_1$trim = 'before'

## reads 2
psqs_2 = extract_data_module(paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt'), 'Per sequence quality scores')
psqs_2$trim = 'after'

comb_psqs = rbind(psqs_1, psqs_2)
comb_psqs$trim = factor(levels = c('before', 'after'), comb_psqs$trim)

p = ggplot(data = comb_psqs) + 
  geom_line(mapping = aes(x = X.Quality, y = Count), color = 'red') + 
  facet_grid(. ~ trim) + 
  scale_x_continuous(name = '\nMean Sequence Qaulity (Phred Score)',
                     limits = c(min(comb_psqs$X.Quality), max(comb_psqs$X.Quality))) +
  scale_y_continuous(name = '') +
  theme(panel.background = element_rect(fill = NA),
        axis.line = element_line(),
        panel.grid.major.y = element_line(color = 'blue', size = 0.1))
p
```

### Per base sequence content

```{r 'Per base sequence content'}
## reads 1
pbsc_1 = extract_data_module(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt'), 'Per base sequence content')
pbsc_1$id = 1:length(pbsc_1$X.Base)

melt_pbsc_1 = melt(pbsc_1, id=c('X.Base', 'id'))
melt_pbsc_1$trim = 'before'


## reads 2
pbsc_2 = extract_data_module(paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt'), 'Per base sequence content')
pbsc_2$id = 1:length(pbsc_2$X.Base)

melt_pbsc_2 = melt(pbsc_2, id=c('X.Base', 'id'))
melt_pbsc_2$trim = 'after'

comb_pbsc = rbind(melt_pbsc_1, melt_pbsc_2)
comb_pbsc$trim = factor(levels = c('before', 'after'), comb_pbsc$trim)

p = ggplot(data = comb_pbsc) +
  geom_line(mapping = aes(x = id, y = value, color = variable)) +
  facet_grid(. ~ trim) +
  xlim(min(comb_pbsc$id), max(comb_pbsc$id)) + 
  ylim(0, 100) +
  xlab('\nPosition in read (bp)') +
  ylab('') +
  scale_color_discrete(name = '') +
  theme_classic()
p
```

### Per sequence GC content

```{r 'Per sequence GC content'}
## reads 1
psGCc_1 = extract_data_module(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt'), 'Per sequence GC content')
psGCc_1$trim = 'before'

## reads 2
psGCc_2 = extract_data_module(paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt'), 'Per sequence GC content')
psGCc_2$trim = 'after'

comb_psGCc = rbind(psGCc_1, psGCc_2)
comb_psGCc$trim = factor(levels = c('before', 'after'), comb_psGCc$trim)

p = ggplot(data = comb_psGCc, aes(x = X.GC.Content, y = Count)) +
  geom_line(color = 'red') +
  facet_grid(. ~ trim) +
  xlab('\nMean Sequence Qaulity (Phred Score)') +
  ylab('') +
  scale_color_discrete(name = '') +
  theme_classic()
p
```


### Per base N content

```{r 'Per base N content'}
## reads 1
pbNc_1 = extract_data_module(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt'), 'Per base N content')
pbNc_1$id = 1:length(pbNc_1$X.Base)
pbNc_1$trim = 'before'

## reads 2
pbNc_2 = extract_data_module(paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt'), 'Per base N content')
pbNc_2$id = 1:length(pbNc_2$X.Base)
pbNc_2$trim = 'after'

comb_pbNc = rbind(pbNc_1, pbNc_2)
comb_pbNc$trim = factor(levels = c('before', 'after'), comb_pbNc$trim)

p = ggplot(data = comb_pbNc, aes(x = id, y = N.Count)) +
  geom_line(color = 'red') +
  scale_x_continuous(breaks = pbNc_2$id, labels = pbNc_2$X.Base) + 
  facet_grid(. ~ trim) +
  ylim(0, 1) + 
  xlab('\nN-Count') +
  ylab('') + 
  theme(axis.text.x = element_text(size = 5),
        axis.line = element_line(),
        panel.background = element_rect(fill = NA))
p
```


### Sequence Length Distribution

```{r 'Sequence Length Distribution'}
## reads 1
sld_1 = extract_data_module(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt'), 'Sequence Length Distribution')
sld_1$id = 1:length(sld_1$X.Length)
sld_1$trim = 'before'

## reads 2
sld_2 = extract_data_module(paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt'), 'Sequence Length Distribution')
sld_2$id = 1:length(sld_2$X.Length)
sld_2$trim = 'after'

comb_sld = rbind(sld_1, sld_2)
comb_sld$trim = factor(levels = c('before', 'after'), comb_sld$trim)

p = ggplot(data = comb_sld, aes(x = id, y = Count)) +
  geom_line(color = 'red') +
  scale_x_continuous(breaks = sld_2$id, labels = sld_2$X.Length) + 
  facet_grid(. ~ trim) +
  xlab('\nSequence Length (bp)') +
  ylab('') + 
  theme(axis.text.x = element_text(size = 5),
        panel.background = element_rect(fill = NA),
        axis.line = element_line(), 
        plot.margin = margin(2,2,2,10) )
p
```

### Sequence Duplication Levels

```{r 'Sequence Duplication Levels'}
## reads 1
sdl_1 = extract_data_module(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt'), 'Sequence Duplication Levels', header = FALSE, comment.char = '#')
names(sdl_1) = c('Duplication_Level', 'Percentage_of_deduplicated', 'Percentage_of_total')
sdl_1$id = 1:length(sdl_1$Duplication_Level)

melt_sdl_1 = melt(sdl_1, id=c('Duplication_Level', 'id'))
melt_sdl_1$trim = 'before'


## reads 2
sdl_2 = extract_data_module(paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt'), 'Sequence Duplication Levels', header = FALSE, comment.char = '#')
names(sdl_2) = c('Duplication_Level', 'Percentage_of_deduplicated', 'Percentage_of_total')
sdl_2$id = 1:length(sdl_2$Duplication_Level)

melt_sdl_2 = melt(sdl_2, id=c('Duplication_Level', 'id'))
melt_sdl_2$trim = 'after'

comb_sdl = rbind(melt_sdl_1, melt_sdl_2)
comb_sdl$trim = factor(levels = c('before', 'after'), comb_sdl$trim)

p = ggplot(data = comb_sdl) +
  geom_line(mapping = aes(x = id, y = value, color = variable)) +
  scale_x_continuous(breaks = sdl_2$id, labels = sdl_2$Duplication_Level) +
  facet_grid(. ~ trim) +
  xlab('\nSequence Duplication Level') +
  ylab('') + 
  scale_color_discrete(name = '') +
  theme(axis.text.x = element_text(size = 5),
        panel.background = element_rect(fill = NA),
        axis.line = element_line(),
        legend.position="top")
p
```


### Overrepresented sequences {.tabset}

#### Before

```{r}
fastqc_data_1 = paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt')
module_name = 'Overrepresented sequences'

overrepresented_seq = extract_data_module(fastqc_data_1, module_name)
colnames(overrepresented_seq) = c('Sequence', 'Count', 'Percentage', 'Possible Source')
DT::datatable(overrepresented_seq)
```

#### After

```{r}
fastqc_data_2 = paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt')
module_name = 'Overrepresented sequences'

overrepresented_seq = extract_data_module(fastqc_data_2, module_name)
colnames(overrepresented_seq) = c('Sequence', 'Count', 'Percentage', 'Possible Source')
DT::datatable(overrepresented_seq)
```


### Adapter Content

```{r 'Adapter Content'}
## reads 1
ac_1 = extract_data_module(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt'), 'Adapter Content')
ac_1$id = 1:length(ac_1$X.Position)

melt_ac_1 = melt(ac_1, id=c('X.Position', 'id'))
melt_ac_1$trim = 'before'

## reads 2
ac_2 = extract_data_module(paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt'), 'Adapter Content')
ac_2$id = 1:length(ac_2$X.Position)

melt_ac_2 = melt(ac_2, id=c('X.Position', 'id'))
melt_ac_2$trim = 'after'

comb_ac = rbind(melt_ac_1, melt_ac_2)
comb_ac$trim = factor(levels = c('before', 'after'), comb_ac$trim)

p = ggplot(data = comb_ac, aes(x = id, y = value, color = variable)) +
  geom_line() +
  facet_grid(. ~ trim) +
  xlim(min(comb_ac$id), max(comb_ac$id)) + 
  ylim(0, 1) +
  xlab('\nPosition in read (bp)') +
  ylab('') +
  scale_color_discrete(name = '') +
  theme(axis.text.x = element_text(size = 5),
        panel.background = element_rect(fill = NA),
        axis.line = element_line())
ggplotly(p)
```

### Kmer Content {.tabset}

#### Before

```{r 'Kmer Content (before)'}
kc_1 = extract_data_module(paste0(opt$X_d, '/read_1_fastqc/fastqc_data.txt'), 'Kmer Content')
DT::datatable(kc_1)
```

#### After
```{r 'Kmer Content (after)'}
kc_2 = extract_data_module(paste0(opt$X_d, '/read_2_fastqc/fastqc_data.txt'), 'Kmer Content')
DT::datatable(kc_2)
```

