---
title: 'Dustmasker Report'
output:
    html_document:
      highlight: pygments
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(error = TRUE, echo = FALSE)
```

```{css echo=FALSE}
# code chunks scrollable
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
```


```{r, echo=FALSE}
# to make the css theme to work, <link></link> tags cannot be added directly 
# as <script></script> tags as below.
# it has to be added using a code chunk with the htmltool functions!!!
css_link = tags$link()
css_link$attribs = list(rel="stylesheet", href="vakata-jstree-3.3.5/dist/themes/default/style.min.css")
css_link
```

```{r, eval=FALSE, echo=FALSE}
# this code chunk is purely for adding comments
# below is to add jQuery and jstree javascripts
```
<script src="https://code.jquery.com/jquery-3.3.1.min.js"></script>
<script src="vakata-jstree-3.3.5/dist/jstree.min.js"></script>

---
# javascript code below is to build the file tree interface
# see this for how to implement opening hyperlink: https://stackoverflow.com/questions/18611317/how-to-get-i-get-leaf-nodes-in-jstree-to-open-their-hyperlink-when-clicked-when
---
<script>
  $(function () {
    // create an instance when the DOM is ready
    $('#jstree').jstree().bind("select_node.jstree", function (e, data) {
     window.open( data.node.a_attr.href, data.node.a_attr.target )
    });
  });
</script>

---
# ADD YOUR DATA ANALYSIS CODE AND MARKUP TEXT BELOW TO EXTEND THIS R MARKDOWN FILE
---


## Run `dustmasker` job

```{bash echo=FALSE}
sh ${TOOL_INSTALL_DIR}/scripts_generator.sh
```

```{r echo=FALSE,warning=FALSE,results='asis'}
# display content of the job-script.sh file.
cat('```bash\n')
cat(readLines(paste0(Sys.getenv('REPORT_FILES_PATH'), '/dustmasker.sh')), sep = '\n')
cat('\n```')
```


**Options explanation:**

<pre>
-in <File_In>
   input file name
   Default = `-'
 -window <Integer>
   DUST window length
   Default = `64'
 -level <Integer>
   DUST level (score threshold for subwindows)
   Default = `20'
 -linker <Integer>
   DUST linker (how close masked intervals should be to get merged together).
   Default = `1'
 -infmt <String>
   input format (possible values: fasta, blastdb)
   Default = `fasta'
 -outfmt <String, `acclist', `fasta', `interval', `maskinfo_asn1_bin',
                  `maskinfo_asn1_text', `maskinfo_xml', `seqloc_asn1_bin',
                  `seqloc_asn1_text', `seqloc_xml'>
   output format
   Default = `interval'
 -parse_seqids
   Parse Seq-ids in FASTA input
   
 -out <File_Out>
   output file name
</pre>