---
title: 'Short reads evaluation with [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/)'
output:
    html_document:
      number_sections: true
      toc: true
      theme: cosmo
      highlight: tango
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = ECHO,
  error = TRUE
)
```


# Fastqc Evaluation

## Evaluation of reads before trimming

```{r}
if ('READS_1' == 'None') {
  stop("No pre-trimming reads provided!")
} else {
  ## run fastqc evaluation
  fastqc_command = paste0('fastqc ') %>% 
  (function(x) {
    ifelse('CONTAMINANTS' != 'None', paste0(x, '-c CONTAMINANTS '), x)
  }) %>% 
  (function(x) {
    ifelse('LIMITS' != 'None', paste0(x, '-l LIMITS '), x)
  }) %>% 
  (function(x) {
    paste0(x, '-o REPORT_DIR ')
  })
  fastqc_command_reads_1 = paste0(fastqc_command, 'READS_1 > /dev/null 2>&1')
  system(fastqc_command_reads_1, intern = TRUE)
  
  # Original html report
  reads_1_base = tail(strsplit('READS_1', '/')[[1]], 1)
  original_html = tags$a(href=paste0(reads_1_base, '_fastqc.html'), paste0('HTML report: ', opt$name_1))
  
  unzip(paste0('REPORT_DIR/', reads_1_base, '_fastqc.zip'), exdir = 'REPORT_DIR')
  reads_1_unzip = paste0('REPORT_DIR/', reads_1_base, '_fastqc/')
  # fastqc_data.txt
  file.copy(paste0(reads_1_unzip, 'fastqc_data.txt'), 'REPORT_DIR/reads_1_fastqc_data.txt')
  fastqc_data = tags$a(href='reads_1_fastqc_data.txt', paste0('fastqc_data.txt: ', opt$name_1))
  # summary.txt
  file.copy(paste0(reads_1_unzip, 'summary.txt'), 'REPORT_DIR/reads_1_summary.txt')
  summary_data = tags$a(href='reads_1_summary.txt', paste0('summary.txt: ', opt$name_1))
  
  tags$ul(
    tags$li(original_html),
    tags$li(fastqc_data),
    tags$li(summary_data)
  )
}
```


## Evaluation of reads after trimming

```{r}
if ('READS_2' == 'None') {
  stop("No pre-trimming reads provided!")
} else {
  ## run fastqc evaluation
  fastqc_command = paste0('fastqc ') %>% 
  (function(x) {
    ifelse('CONTAMINANTS' != 'None', paste0(x, '-c CONTAMINANTS '), x)
  }) %>% 
  (function(x) {
    ifelse('LIMITS' != 'None', paste0(x, '-l LIMITS '), x)
  }) %>% 
  (function(x) {
    paste0(x, '-o REPORT_DIR ')
  })
  fastqc_command_reads_2 = paste0(fastqc_command, 'READS_2 > /dev/null 2>&1')
  system(fastqc_command_reads_2, intern = TRUE)
  
  # Original html report
  reads_2_base = tail(strsplit('READS_2', '/')[[1]], 1)
  original_html = tags$a(href=paste0(reads_2_base, '_fastqc.html'), paste0('HTML report: ', opt$name_2))
  
  unzip(paste0('REPORT_DIR/', reads_2_base, '_fastqc.zip'), exdir = 'REPORT_DIR')
  reads_2_unzip = paste0('REPORT_DIR/', reads_2_base, '_fastqc/')
  # fastqc_data.txt
  file.copy(paste0(reads_2_unzip, 'fastqc_data.txt'), 'REPORT_DIR/reads_2_fastqc_data.txt')
  fastqc_data = tags$a(href='reads_2_fastqc_data.txt', paste0('fastqc_data.txt: ', opt$name_2))
  # summary.txt
  file.copy(paste0(reads_2_unzip, 'summary.txt'), 'REPORT_DIR/reads_2_summary.txt')
  summary_data = tags$a(href='reads_2_summary.txt', paste0('summary.txt: ', opt$name_2))
  
  tags$ul(
    tags$li(original_html),
    tags$li(fastqc_data),
    tags$li(summary_data)
  )
}
```



# Fastqc output visualization

## Overview

```{r}
reads_1_summary = read.csv('REPORT_DIR/reads_1_summary.txt', header = FALSE, sep = '\t')[, 2:1]
reads_2_summary = read.csv('REPORT_DIR/reads_2_summary.txt', header = FALSE, sep = '\t')[, 1]
combined_summary = cbind(reads_1_summary, reads_2_summary)
names(combined_summary) = c('MODULE', paste0(opt$name_1, '(before)'), paste0(opt$name_2, '(after)'))
combined_summary[combined_summary == 'FAIL'] = 'FAIL (X)'
combined_summary[combined_summary == 'WARN'] = 'WARN (!)'
knitr::kable(combined_summary)
```

## Visualization by data module {.tabset}

* Define a function to extract outputs for each module from fastqc output

```{r 'function definition'}
extract_data_module = function(fastqc_data, module_name, header = TRUE, comment.char = "") {
  f = readLines(fastqc_data)
  start_line = grep(module_name, f)
  end_module_lines = grep('END_MODULE', f)
  end_line = end_module_lines[which(end_module_lines > start_line)[1]]
  module_data = f[(start_line+1):(end_line-1)]
  writeLines(module_data, 'temp.txt')
  read.csv('temp.txt', sep = '\t', header = header, comment.char = comment.char)
}
```

### Per base sequence quality

```{r 'per base sequence quality', fig.width=10}
## reads 1
pbsq_1 = extract_data_module('REPORT_DIR/reads_1_fastqc_data.txt', 'Per base sequence quality')
pbsq_1$id = 1:length(pbsq_1$X.Base)
pbsq_1$trim = 'before'

## reads 2
pbsq_2 = extract_data_module('REPORT_DIR/reads_2_fastqc_data.txt', 'Per base sequence quality')
pbsq_2$id = 1:length(pbsq_2$X.Base)
pbsq_2$trim = 'after'

comb_pbsq = rbind(pbsq_1, pbsq_2)
comb_pbsq$trim = factor(levels = c('before', 'after'), comb_pbsq$trim)

p = ggplot(data = comb_pbsq) +
  geom_boxplot(mapping = aes(x = id, 
                             lower = Lower.Quartile, 
                             upper = Upper.Quartile, 
                             middle = Median, 
                             ymin = X10th.Percentile, 
                             ymax = X90th.Percentile,
                             fill = "yellow"),
               stat = 'identity') +
  geom_line(mapping = aes(x = id, y = Mean, color = "red")) +
  scale_x_continuous(breaks = pbsq_2$id, labels = pbsq_2$X.Base) +
  scale_fill_identity() +
  scale_color_identity() + 
  ylim(0, max(comb_pbsq$Upper.Quartile) + 5) +
  xlab('Position in read (bp)') +
  facet_grid(. ~ trim) +
  theme(axis.text.x = element_text(angle=45))
p

```

### Per tile sequence quality

```{r 'per tile sequence quality', fig.width=10}
## check if 'per tile sequence quality' module exits or not
check_ptsq = grep('Per tile sequence quality', readLines('REPORT_DIR/reads_1_fastqc_data.txt'))
if (length(check_ptsq) > 0) {
    ## reads 1
  ptsq_1 = extract_data_module('REPORT_DIR/reads_1_fastqc_data.txt', 'Per tile sequence quality')
  ptsq_1$trim = 'before'
  
  ## reads 2
  ptsq_2 = extract_data_module('REPORT_DIR/reads_2_fastqc_data.txt', 'Per tile sequence quality')
  ptsq_2$trim = 'after'
  
  comb_ptsq = rbind(ptsq_1, ptsq_2)
  comb_ptsq$trim = factor(levels = c('before', 'after'), comb_ptsq$trim)
  comb_ptsq$Base = factor(levels = unique(comb_ptsq$Base), comb_ptsq$Base)
  
  # convert integers to charaters
  comb_ptsq$Tile = as.character(comb_ptsq$X.Tile)
  
  p = ggplot(data = comb_ptsq, aes(x = Base, y = Tile, fill = Mean)) +
    geom_raster() + 
    facet_grid(. ~ trim) + 
    xlab('Position in read (bp)') + 
    ylab('') +
    theme(axis.text.x = element_text(angle=45))
  ggplotly(p)
} else {
  print('No "per tile sequence quality" data')
}


```

### Per sequence quality score

```{r 'Per sequence quality score', fig.width=10}
## reads 1
psqs_1 = extract_data_module('REPORT_DIR/reads_1_fastqc_data.txt', 'Per sequence quality scores')
psqs_1$trim = 'before'

## reads 2
psqs_2 = extract_data_module('REPORT_DIR/reads_2_fastqc_data.txt', 'Per sequence quality scores')
psqs_2$trim = 'after'

comb_psqs = rbind(psqs_1, psqs_2)
comb_psqs$trim = factor(levels = c('before', 'after'), comb_psqs$trim)

p = ggplot(data = comb_psqs, aes(x = X.Quality, y = Count)) + 
  geom_line(color = 'red') + 
  facet_grid(. ~ trim) + 
  xlim(min(comb_psqs$X.Quality), max(comb_psqs$X.Quality)) + 
  xlab('Mean Sequence Qaulity (Phred Score)') + 
  ylab('')
ggplotly(p)
```


### Per base sequence content

```{r 'Per base sequence content', fig.width=10}
## reads 1
pbsc_1 = extract_data_module('REPORT_DIR/reads_1_fastqc_data.txt', 'Per base sequence content')
pbsc_1$id = 1:length(pbsc_1$X.Base)

melt_pbsc_1 = melt(pbsc_1, id=c('X.Base', 'id'))
melt_pbsc_1$trim = 'before'


## reads 2
pbsc_2 = extract_data_module('REPORT_DIR/reads_2_fastqc_data.txt', 'Per base sequence content')
pbsc_2$id = 1:length(pbsc_2$X.Base)

melt_pbsc_2 = melt(pbsc_2, id=c('X.Base', 'id'))
melt_pbsc_2$trim = 'after'

comb_pbsc = rbind(melt_pbsc_1, melt_pbsc_2)
comb_pbsc$trim = factor(levels = c('before', 'after'), comb_pbsc$trim)

p = ggplot(data = comb_pbsc, aes(x = id, y = value, color = variable)) +
  geom_line() +
  facet_grid(. ~ trim) +
  xlim(min(comb_pbsc$id), max(comb_pbsc$id)) + 
  ylim(0, 100) +
  xlab('Position in read (bp)') +
  ylab('')
ggplotly(p)
```

### Per sequence GC content

```{r 'Per sequence GC content', fig.width=10}
## reads 1
psGCc_1 = extract_data_module('REPORT_DIR/reads_1_fastqc_data.txt', 'Per sequence GC content')
psGCc_1$trim = 'before'

## reads 2
psGCc_2 = extract_data_module('REPORT_DIR/reads_2_fastqc_data.txt', 'Per sequence GC content')
psGCc_2$trim = 'after'

comb_psGCc = rbind(psGCc_1, psGCc_2)
comb_psGCc$trim = factor(levels = c('before', 'after'), comb_psGCc$trim)

p = ggplot(data = comb_psGCc, aes(x = X.GC.Content, y = Count)) +
  geom_line(color = 'red') +
  facet_grid(. ~ trim) +
  xlab('Mean Sequence Qaulity (Phred Score)') +
  ylab('')
ggplotly(p)
```


### Per base N content

```{r 'Per base N content', fig.width=10}
## reads 1
pbNc_1 = extract_data_module('REPORT_DIR/reads_1_fastqc_data.txt', 'Per base N content')
pbNc_1$id = 1:length(pbNc_1$X.Base)
pbNc_1$trim = 'before'

## reads 2
pbNc_2 = extract_data_module('REPORT_DIR/reads_2_fastqc_data.txt', 'Per base N content')
pbNc_2$id = 1:length(pbNc_2$X.Base)
pbNc_2$trim = 'after'

comb_pbNc = rbind(pbNc_1, pbNc_2)
comb_pbNc$trim = factor(levels = c('before', 'after'), comb_pbNc$trim)

p = ggplot(data = comb_pbNc, aes(x = id, y = N.Count)) +
  geom_line(color = 'red') +
  scale_x_continuous(breaks = pbNc_2$id, labels = pbNc_2$X.Base) + 
  facet_grid(. ~ trim) +
  ylim(0, 1) + 
  xlab('N-Count') +
  ylab('') + 
  theme(axis.text.x = element_text(angle=45))
ggplotly(p)
```


### Sequence Length Distribution

```{r 'Sequence Length Distribution', fig.width=10}
## reads 1
sld_1 = extract_data_module('REPORT_DIR/reads_1_fastqc_data.txt', 'Sequence Length Distribution')
sld_1$id = 1:length(sld_1$X.Length)
sld_1$trim = 'before'

## reads 2
sld_2 = extract_data_module('REPORT_DIR/reads_2_fastqc_data.txt', 'Sequence Length Distribution')
sld_2$id = 1:length(sld_2$X.Length)
sld_2$trim = 'after'

comb_sld = rbind(sld_1, sld_2)
comb_sld$trim = factor(levels = c('before', 'after'), comb_sld$trim)

p = ggplot(data = comb_sld, aes(x = id, y = Count)) +
  geom_line(color = 'red') +
  scale_x_continuous(breaks = sld_2$id, labels = sld_2$X.Length) + 
  facet_grid(. ~ trim) +
  xlab('Sequence Length (bp)') +
  ylab('') + 
  theme(axis.text.x = element_text(angle=45))
ggplotly(p)
```

### Sequence Duplication Levels

```{r 'Sequence Duplication Levels', fig.width=10}
## reads 1
sdl_1 = extract_data_module('REPORT_DIR/reads_1_fastqc_data.txt', 'Sequence Duplication Levels', header = FALSE, comment.char = '#')
names(sdl_1) = c('Duplication_Level', 'Percentage_of_deduplicated', 'Percentage_of_total')
sdl_1$id = 1:length(sdl_1$Duplication_Level)

melt_sdl_1 = melt(sdl_1, id=c('Duplication_Level', 'id'))
melt_sdl_1$trim = 'before'


## reads 2
sdl_2 = extract_data_module('REPORT_DIR/reads_2_fastqc_data.txt', 'Sequence Duplication Levels', header = FALSE, comment.char = '#')
names(sdl_2) = c('Duplication_Level', 'Percentage_of_deduplicated', 'Percentage_of_total')
sdl_2$id = 1:length(sdl_2$Duplication_Level)

melt_sdl_2 = melt(sdl_2, id=c('Duplication_Level', 'id'))
melt_sdl_2$trim = 'after'

comb_sdl = rbind(melt_sdl_1, melt_sdl_2)
comb_sdl$trim = factor(levels = c('before', 'after'), comb_sdl$trim)

p = ggplot(data = comb_sdl, aes(x = id, y = value, color = variable)) +
  geom_line() +
  scale_x_continuous(breaks = sdl_2$id, labels = sdl_2$Duplication_Level) +
  facet_grid(. ~ trim) +
  xlab('Sequence Duplication Level') +
  ylab('') + 
  theme(axis.text.x = element_text(angle=45))
ggplotly(p)
```

### Adapter Content

```{r 'Adapter Content', fig.width=10}
## reads 1
ac_1 = extract_data_module('REPORT_DIR/reads_1_fastqc_data.txt', 'Adapter Content')
ac_1$id = 1:length(ac_1$X.Position)

melt_ac_1 = melt(ac_1, id=c('X.Position', 'id'))
melt_ac_1$trim = 'before'

## reads 2
ac_2 = extract_data_module('REPORT_DIR/reads_2_fastqc_data.txt', 'Adapter Content')
ac_2$id = 1:length(ac_2$X.Position)

melt_ac_2 = melt(ac_2, id=c('X.Position', 'id'))
melt_ac_2$trim = 'after'

comb_ac = rbind(melt_ac_1, melt_ac_2)
comb_ac$trim = factor(levels = c('before', 'after'), comb_ac$trim)

p = ggplot(data = comb_ac, aes(x = id, y = value, color = variable)) +
  geom_line() +
  facet_grid(. ~ trim) +
  xlim(min(comb_ac$id), max(comb_ac$id)) + 
  ylim(0, 1) +
  xlab('Position in read (bp)') +
  ylab('')
ggplotly(p)
```

### Kmer Content {.tabset}

#### Before

```{r 'Kmer Content (before)', fig.width=10}
kc_1 = extract_data_module('REPORT_DIR/reads_1_fastqc_data.txt', 'Kmer Content')
knitr::kable(kc_1)
```

#### After
```{r 'Kmer Content (after)', fig.width=10}
kc_2 = extract_data_module('REPORT_DIR/reads_2_fastqc_data.txt', 'Kmer Content')
knitr::kable(kc_2)
```


# Session Info

```{r 'session info'}
sessionInfo()
```

# References

* Bioinformatics, Babraham (2014). FastQC.

* Allaire, J and Cheng, Joe and Xie, Yihui and McPherson, Jonathan and Chang, Winston and Allen, Jeff and Wickham, Hadley and Atkins, Aron and Hyndman, Rob (2016). rmarkdown: Dynamic Documents for R, 2016. In R package version 0.9, 6.

* Xie, Yihui (2015). Dynamic Documents with R and knitr, CRC Press, Vol.29.

* Carson Sievert and Chris Parmer and Toby Hocking and Scott Chamberlain and Karthik Ram and Marianne Corvellec and Pedro Despouy (2017). plotly: Create Interactive Web Graphics via 'plotly.js'. R package version 4.6.0. [Link]

* Wickham, H. (2016). ggplot2: elegant graphics for data analysis. Springer. Chicago
